Persistence in functional topology and a correction to a theorem of Morse

### 2021-07-26 connectivity.tex
Pre-release version

For the rest of this section, we will assume that $\H$ is a homology theory that has either the open or the compact Mayer-Vietoris property and for which there is $n_0$ such that $\H_{n}$ is 0 for all $n \leq n_0$.
Note that this includes singular homology with field coefficients, which, like any homology theory in the sense of the Eilenberg-Steenrod axioms \cite[Section I]{MR0050886}, has the open Mayer-Vietoris property, and it also includes \v{C}ech homology with field coefficients, which has the closed Mayer-Vietoris property.

@@

Clearly, any strongly $\HLC$ sublevel set filtration is also $\HLC$.
If the filtration is induced by a continuous function, the converse also holds as the following theorem shows.

>>>>

For the rest of this section, we will assume that $\H$ is a homology theory that has either the open or the compact Mayer-Vietoris property and for which there is $n_0$ such that $\H_{n}$ is 0 for all $n \leq n_0$.
Note that this includes singular homology with field coefficients, which, like any homology theory in the sense of the Eilenberg-Steenrod axioms \cite[Section I]{MR0050886}, has the open Mayer-Vietoris property, and it also includes \v{C}ech homology with field coefficients, which has the compact Mayer-Vietoris property.

@@

Clearly, any strongly $\HLC$ sublevel set filtration is also $\HLC$:
while the $\HLC$ property merely requires the existence of an index $s \in (f(x),t)$ satisfying the $\HS$ condition, the \emph{strong} $\HLC$ property requires the $\HS$ condition to hold for any $s \in (f(x),t)$.
If the filtration is induced by a continuous function, the converse also holds as the following theorem shows.

### 2021-07-26 inequalities.tex
Pre-release version

This approach of counting homological changes instead of critical points is also what Morse used in the non-smooth setting of functional topology.
To keep track of the number of $d$-dimensional homological events at filtration value $t$ that persist for at least time $e$ but not indefinitely, \citet{Morse.1940} defined the $(d, t, e)$-\textit{cap numbers} of a filtration, which, when the associated persistence module has a barcode and is upper semi-continuous\footnote{A persistence module $M$ is \emph{upper semi-continuous} if $M_{t} \to \lim_{u > t} M_{u}$ is an isomorphism for all~$t$.}, correspond to the number of finite bars with length greater than $e$ in the $d\th$ barcode with left endpoint $t$, plus the number of finite bars with length greater than $e$ in the $(d-1)\th$ barcode with right endpoint $t$ (killed by something $d$-dimensional).

Working entirely in the algebraic setting, we will now fix a graded q-tame persistence module $M$.
Of course, one may think of $M$ as the persistent homology of a q-tame filtration, but $M$ could for example also arise as the filtered Floer homology of some Hamiltonian on a symplectic manifold.
By \cref{thm:q-tame modules have barcodes}, $M$ has a persistence diagram in every degree $d$ with multiplicity function denoted $\mathfrak{m}_d \colon \multiplicityDomain \to \N$.
In analogy to Morse's definitions, we may define the $(d, t, e)$-\textit{cap number} of our graded q-tame persistence module $M$ in terms of its persistence diagram as
\begin{equation*}
m_{d}^{> e}(t) =
\sum_{\substack{p \in \R \\ t - p > e}} \mathfrak{m}_{d-1}(p, t) +
\sum_{\substack{q \in \R \\ q - t > e}} \mathfrak{m}_d(t, q).
\end{equation*}
Note that finiteness of these quantities is ensured by the q-tameness of $M$ and the use of a non-zero $e$ bounding below the persistence of the considered features.
To see the necessity of this second condition, consider the q-tame persistence module given by the infinite product $\prod_{n \in \N_{> 0}} C([0,1/n))$.

Whenever the sums below are well-defined, we also consider the \emph{$(d,e)$-cap numbers} 
\begin{equation} \label{e:cap numbers}
m_{d}^{> e } =
\sum_{t} m_{d}^{> e}(t) \ =
\sum_{\substack{(p,q) \in \multiplicityDomain \\ q - p > e \\ q \neq \infty}} \mathfrak{m}_{d-1}(p,q) \ +
\sum_{\substack{(p,q) \in \multiplicityDomain \\ q - p > e \\ p \neq -\infty}}\mathfrak{m}_d(p,q).
\end{equation} 
They are well-defined for example if $M$ is the persistent homology of the sublevel set filtration of a bounded function.
This can be deduced from the following general result.

\begin{thm}
    Assume that $N$ is a q-tame persistence module that is initially and eventually constant, i.e., that there exist $t_0$ and $t_1$ with $N_s = N_{t_0}$ for all $s \leq t_0$ and $N_u = N_{t_1}$ for all $u \geq t_1$, and let $\mathfrak{n}$ be the multiplicity function of the persistence diagram of $N$.
    Then for each $e>0$ we have 
    \begin{equation*}
    \sum_{ \substack{ (p,q) \in \multiplicityDomain \\ q-p > e } } \mathfrak{n} (p,q) < \infty.
    \end{equation*}
\end{thm}

\begin{proof}
    We split the sum whose finiteness we want to show in two parts
    \begin{equation*}
    \sum_{ \substack{ (p,q) \in \multiplicityDomain \\ q-p > e } } \mathfrak{n} (p,q)
    \ = \!\!
    \sum_{ \substack{ (p,q) \in \multiplicityDomain \setminus T \\ q-p > e } } \mathfrak{n} (p,q)
    \ \ +
    \sum_{ \substack{ (p,q) \in T \\ q-p > e } } \mathfrak{n} (p,q)
    \end{equation*}
    and show for each summand separately that it is finite.
    Here, $T$ denotes the triangle
    \begin{equation*}
    T = \{(p,q) \in \multiplicityDomain \mid t_0 \leq p < q \leq t_1\}.
    \end{equation*}

@@

    For the second summand, note that  
    \begin{equation*}
    \sum_{ \substack{ (p,q) \in T \\ q-p > e } } \mathfrak{n} (p,q)
    \ \, \leq \!\!
    \sum_{(p,q) \in T^{\geq e}} \mathfrak{n} (p,q),
    \end{equation*}
    where $T^{\geq e}$ is the smaller triangle 
    \begin{equation*}
    T^{\geq e} = \{(p,q) \in T \mid q-p \geq e\}.
    \end{equation*}
    Thus, in order to prove the theorem, it suffices to show that we have 
    \begin{equation*}
    \sum_{(p,q) \in T^{\geq e}} \mathfrak{n} (p,q) 
    \ < \ 
    \infty.
    \end{equation*}
    To do this, we consider open quadrants 
    \begin{equation*}
    Q(x, y) = \{ (p, q) \in \R^2 \mid p < x \text{ and } y < q \}.
    \end{equation*}
    Covering the compact set $T^{\geq e}$ by the open quadrants $Q \left(x, x + \frac{e}{2} \right)$ for $x \in \R$, we may choose a finite subcover given by, say, $x_1,\dots, x_n$.
	We obtain
    \begin{equation*}
    \sum_{(p,q) \in T^{\geq e}} \mathfrak{n} (p,q) 
    \ \leq \
    \sum_{i=1}^n \sum_{\substack{(p, q) \in \\ Q (x_i, x_i + \frac{e}{2})}} \mathfrak{n}(p,q).
    \end{equation*}
    Each of sums $\sum_{(p,q) \in Q \left(x_i, x_i + \frac{e}{2} \right)} \mathfrak{n}(p,q)$ over the quadrants $Q \left(x_i, x_i + \frac{e}{2} \right)$ is finite since $N$ is q-tame (which is where the name q-tame or \emph{quadrant}-tame comes from, see \cite[Section 3.8]{Chazal.2016a}).
\end{proof}

@@

\begin{thm}\label{t:inequalities}
	Let $M$ be a graded q-tame persistence module with cap numbers $m_{d}^{> e }$ and essential dimensions $p_{d}$ finite for all $d$ and $e$.
	If $\mathfrak{m}_d(-\infty, p) = 0$ for all $p \in \R \cup \{\infty\}$ and all $d$, then we have Morse inequalities
    \begin{equation} \label{e:morse inequalities}
    \sum_{d=0}^n \ (-1)^{n-d} (m^{>e}_{d} - p_{d}) \ \geq\  0
    \end{equation}
    for any dimension $n$ and any $e > 0$.
\end{thm}

\begin{proof}
    Using a telescopic sum argument, the claimed inequalities can be seen to be equivalent to the existence of a sequence $(\nu_d)_d$ of non-negative integers with $m^{>e}_{d} - p_{d} = \nu_{d-1} + \nu_{d}$.
    Since we assume $\mathfrak{m}_d(-\infty, p) = 0$ for all $p$, we have
    \begin{align*}
    m^{>e}_{d}
    \ &=
    \sum_{\substack{(p,q) \in \multiplicityDomain \\ q - p > e \\ q \neq \infty}} \mathfrak{m}_{d-1}(p,q)
    \ +
    \sum_{\substack{(p,q) \in \multiplicityDomain \\ q - p > e \\ p \neq -\infty}}\mathfrak{m}_d(p,q)
    \\
    &=
    \sum_{\substack{(p,q) \in \R^2 \\ q - p > e }} \mathfrak{m}_{d-1}(p,q)
    \ +
    \sum_{\substack{(p,q) \in \multiplicityDomain \\ q - p > e \\ p \neq -\infty}}\mathfrak{m}_d(p,q)
    \end{align*}
    and
    \begin{equation*}
    p_{d}
    \ = \!\!
    \sum_{p \in \R \cup \{-\infty\}} \mathfrak{m}_d(p,\infty)
    \ = \
    \sum_{p \in \R} \mathfrak{m}_d(p,\infty),
    \end{equation*}
    which yields
    \begin{align*}
    m^{>e}_{d} - p_{d} \
    &= 
    \sum_{\substack{(p,q) \in \R^2 \\ q - p > e }} \mathfrak{m}_{d-1}(p,q)
    \ +
    \sum_{\substack{(p,q) \in \multiplicityDomain \\ q - p > e \\ p \neq -\infty}}\mathfrak{m}_d(p,q)
    \ - \
    \sum_{p \in \R} \mathfrak{m}_d(p,\infty)
    \\
    &=
    \sum_{\substack{(p,q) \in \R^2 \\ q - p > e }} \mathfrak{m}_{d-1}(p,q)
    \ + \!
    \sum_{\substack{(p,q) \in \R^2 \\ q - p > e }}\mathfrak{m}_d(p,q),
    \end{align*}
    so the sequence given by $\nu_d = \sum_{\substack{(p,q) \in \R^2 \\ q - p > e }}\mathfrak{m}_d(p,q)$ satisfies $m^{>e}_{d} - p_{d} = \nu_{d-1} + \nu_{d}$ as desired.
\end{proof}

As we have remarked, the finiteness assumptions are satisfied if $M$ is initially and eventually constant, so as a special case, the theorem yields Morse inequalities for any bounded real-valued function whose sublevel set filtration has q-tame persistent homology, which includes classical Morse functions $f$ on closed smooth manifolds $X$.
As we have outlined while motivating the definition of the cap numbers, our inequalities as presented in \eqref{e:morse inequalities} agree with the classical inequalities \eqref{equ:morse_inequalities} because, in this setting, $\beta_d(X) = p_d$ and $m_d^{>e} = \crit_d(f)$ for $e > 0$ smaller than the minimal difference between any two critical values of $f$.

The finiteness conditions of our theorem are also satisfied for the functionals that Morse considered in his work on functional topology, with q-tameness being guaranteed by \cref{t:strong local connectedness implies q-tameness} instead of the insufficient condition Morse proposed.
As we have also mentioned before, our inequalities then agree with Morse's historical version from functional topology as presented in \cite{Morse.1940} if the persistent homology of the sublevel set filtration is not only q-tame but also upper semi-continuous, which is satisfied for compact sublevel set filtrations if one uses \v{C}ech homology, as is the case in Morse's work.
Note, however, that in contrast to Morse's version, our inequalities also work without the upper semi-continuity assumption, removing the need for this unusual choice of homology theory.

\begin{rem}
	In addition to the formulation of the inequalities in terms of cap numbers, Morse also proposes a generalized version of critical points, which he calls homotopically critical and which formalizes the idea of a point being critical if it produces topological changes in the sublevel set filtration.
	This notion is employed by Morse and Tompkins in their study of minimal surfaces \cite{Morse.1939}, which we will partly review in \cref{s:surfaces} and for which a thorough historical account may be found in \cite[Section II.6]{Struwe.1988}.
	The usefulness of this notion might be limited though, because for example in the context of Floer theory, critical points of the action functional corresponding to a Hamiltonian usually have infinite index and thus do not produce changes in the topology.
	In this setting, our approach of formulating the inequalities purely in algebraic terms might be more suitable, because while these critical points are not topologically visible, they do correspond to features in the persistence diagram of the filtered Floer homology, to which our inequalities may also be applied.
\end{rem}

>>>>

This approach of counting homological changes instead of critical points is also what Morse used in the non-smooth setting of functional topology.
To keep track of the number of $d$-dimensional homological events at filtration value $t$ that persist for at least time $\epsilon > 0$ but not indefinitely, \citet{Morse.1940} defined the $(d, t, \epsilon)$-\textit{cap numbers} of a filtration, which, when the associated persistence module has a barcode and is continuous from above, correspond to the number of finite bars with length greater than $\epsilon$ in the $d\th$ barcode with left endpoint $t$, plus the number of finite bars with length greater than $\epsilon$ in the $(d-1)\th$ barcode with right endpoint $t$ (killed by something $d$-dimensional).

Working entirely in the algebraic setting, we will now fix a graded q-tame persistence module $M$.
Of course, one may think of $M$ as the persistent homology of a q-tame filtration, but $M$ could for example also arise as the filtered Floer homology of some Hamiltonian on a symplectic manifold.
By \cref{thm:q-tame modules have barcodes}, $M$ has a persistence diagram in every degree $d$ with multiplicity function denoted $\mathfrak{m}_d \colon \multiplicityDomain \to \N$.
In analogy to Morse's definitions, we may define the $(d, t, \epsilon)$-\textit{cap number} of our graded q-tame persistence module $M$ in terms of its persistence diagram as
\begin{equation*}
m_{d}^{\epsilon}(t) =
\sum_{\substack{p \in \R \\ t - p > \epsilon}} \mathfrak{m}_{d-1}(p, t) +
\sum_{\substack{q \in \R \\ q - t > \epsilon}} \mathfrak{m}_d(t, q).
\end{equation*}
Note that finiteness of these quantities is ensured by the q-tameness of $M$ and the use of a non-zero $e$ bounding below the persistence of the considered features.
To see the necessity of this second condition, consider the q-tame persistence module given by the infinite product $\prod_{n \in \N_{> 0}} C([0,1/n))$.

Whenever the sums below are well-defined, we also consider the \emph{$(d,\epsilon)$-cap numbers} 
\begin{equation} \label{e:cap numbers}
m_{d}^{\epsilon} =
\sum_{t} m_{d}^{\epsilon}(t) \ =
\sum_{\substack{(p,q) \in \multiplicityDomain \\ q - p > \epsilon \\ q \neq \infty}} \mathfrak{m}_{d-1}(p,q) \ +
\sum_{\substack{(p,q) \in \multiplicityDomain \\ q - p > \epsilon \\ p \neq -\infty}}\mathfrak{m}_d(p,q).
\end{equation} 
They are well-defined, for example, if $M$ is the persistent homology of the sublevel set filtration of a bounded function.
This can be deduced from the following general result.

\begin{thm}
    Assume that $N$ is a q-tame persistence module that is initially and eventually constant, i.e., that there exist $t_0$ and $t_1$ with $N_s = N_{t_0}$ for all $s \leq t_0$ and $N_u = N_{t_1}$ for all $u \geq t_1$, and let $\mathfrak{n}$ be the multiplicity function of the persistence diagram of $N$.
    Then for each $\epsilon > 0$ we have 
    \begin{equation*}
    \sum_{ \substack{ (p,q) \in \multiplicityDomain \\ q-p > \epsilon } } \mathfrak{n} (p,q) < \infty.
    \end{equation*}
\end{thm}

\begin{proof}
    We split the sum whose finiteness we want to show in two parts
    \begin{equation*}
    \sum_{ \substack{ (p,q) \in \multiplicityDomain \\ q-p > \epsilon } } \mathfrak{n} (p,q)
    \ = \!\!
    \sum_{ \substack{ (p,q) \in \multiplicityDomain \setminus T \\ q-p > \epsilon } } \mathfrak{n} (p,q)
    \ +
    \sum_{ \substack{ (p,q) \in T \\ q-p > \epsilon } } \mathfrak{n} (p,q)
    \end{equation*}
    and show for each summand separately that it is finite.
    Here, $T$ denotes the triangle
    \begin{equation*}
    T = \{(p,q) \in \multiplicityDomain \mid t_0 \leq p < q \leq t_1\}.
    \end{equation*}

@@

    For the second summand, note that  
    \begin{equation*}
    \sum_{ \substack{ (p,q) \in T \\ q-p > \epsilon } } \mathfrak{n} (p,q)
    \ \, \leq \!\!
    \sum_{(p,q) \in T^{\geq e}} \mathfrak{n} (p,q),
    \end{equation*}
    where $T^{\geq \epsilon}$ is the smaller triangle 
    \begin{equation*}
    T^{\epsilon} = \{(p,q) \in T \mid q-p \geq \epsilon\}.
    \end{equation*}
    Thus, in order to prove the theorem, it suffices to show that we have 
    \begin{equation*}
    \sum_{(p,q) \in T^{e}} \mathfrak{n} (p,q) 
    \ < \ 
    \infty.
    \end{equation*}
    To do this, we consider open quadrants 
    \begin{equation*}
    Q(x, y) = \{ (p, q) \in \R^2 \mid p < x \text{ and } y < q \}.
    \end{equation*}
    Covering the compact set $T^{\epsilon}$ by the open quadrants $Q \left(x, x + \frac{\epsilon}{2} \right)$ for $x \in \R$, we may choose a finite subcover given by, say, $x_1,\dots, x_n$.
	We obtain
    \begin{equation*}
    \sum_{(p,q) \in T^{\geq e}} \mathfrak{n} (p,q) 
    \ \leq \
    \sum_{i=1}^n \sum_{\substack{(p, q) \in \\ Q (x_i, x_i + \frac{\epsilon}{2})}} \mathfrak{n}(p,q).
    \end{equation*}
    Each of the sums $\sum_{(p,q) \in Q \left(x_i, x_i + \frac{\epsilon}{2} \right)} \mathfrak{n}(p,q)$ over the quadrants $Q \left(x_i, x_i + \frac{\epsilon}{2} \right)$ is finite since $N$ is q-tame (which is where the name q-tame or \emph{quadrant}-tame comes from, see \cite[Section 3.8]{Chazal.2016a}).
\end{proof}

@@

\begin{thm}\label{t:inequalities}
	Let $M$ be a graded q-tame persistence module with cap numbers $m_{d}^{\epsilon}$ and essential dimensions $p_{d}$ finite for all $d$ and $\epsilon$.
	If $\mathfrak{m}_d(-\infty, p) = 0$ for all $p \in \R \cup \{\infty\}$ and all $d$, then we have Morse inequalities
    \begin{equation} \label{e:morse inequalities}
    \sum_{d=0}^n \ (-1)^{n-d} (m^{\epsilon}_{d} - p_{d}) \ \geq\  0
    \end{equation}
    for any dimension $n$ and any $\epsilon > 0$.
\end{thm}

\begin{proof}
    Using a telescopic sum argument, the claimed inequalities can be seen to be equivalent to the existence of a sequence $(\nu_d)_d$ of non-negative integers with $m^{\epsilon}_{d} - p_{d} = \nu_{d-1} + \nu_{d}$.
    Since we assume $\mathfrak{m}_d(-\infty, p) = 0$ for all $p$, we have
    \begin{align*}
    m^{\epsilon}_{d}
    \ &=
    \sum_{\substack{(p,q) \in \multiplicityDomain \\ q - p > \epsilon \\ q \neq \infty}} \mathfrak{m}_{d-1}(p,q)
    \ +
    \sum_{\substack{(p,q) \in \multiplicityDomain \\ q - p > \epsilon \\ p \neq -\infty}}\mathfrak{m}_d(p,q)
    \\
    &=
    \sum_{\substack{(p,q) \in \R^2 \\ q - p > \epsilon }} \mathfrak{m}_{d-1}(p,q)
    \ +
    \sum_{\substack{(p,q) \in \multiplicityDomain \\ q - p > \epsilon \\ p \neq -\infty}}\mathfrak{m}_d(p,q)
    \end{align*}
    and
    \begin{equation*}
    p_{d}
    \ = \!\!
    \sum_{p \in \R \cup \{-\infty\}} \mathfrak{m}_d(p,\infty)
    \ = \
    \sum_{p \in \R} \mathfrak{m}_d(p,\infty),
    \end{equation*}
    which yields
    \begin{align*}
    m^{\epsilon}_{d} - p_{d} \
    &= 
    \sum_{\substack{(p,q) \in \R^2 \\ q - p > \epsilon }} \mathfrak{m}_{d-1}(p,q)
    \ +
    \sum_{\substack{(p,q) \in \multiplicityDomain \\ q - p > \epsilon \\ p \neq -\infty}}\mathfrak{m}_d(p,q)
    \ - \
    \sum_{p \in \R} \mathfrak{m}_d(p,\infty)
    \\
    &=
    \sum_{\substack{(p,q) \in \R^2 \\ q - p > \epsilon }} \mathfrak{m}_{d-1}(p,q)
    \ + \!
    \sum_{\substack{(p,q) \in \R^2 \\ q - p > \epsilon }}\mathfrak{m}_d(p,q),
    \end{align*}
    so the sequence given by $\nu_d = \sum_{\substack{(p,q) \in \R^2 \\ q - p > \epsilon }}\mathfrak{m}_d(p,q)$ satisfies $m^{\epsilon}_{d} - p_{d} = \nu_{d-1} + \nu_{d}$ as desired.
\end{proof}

As we have remarked, the finiteness assumptions are satisfied if $M$ is initially and eventually constant, so as a special case, the theorem yields Morse inequalities for any bounded real-valued function whose sublevel set filtration has q-tame persistent homology, which includes classical Morse functions $f$ on closed smooth manifolds $X$.
As we have outlined while motivating the definition of the cap numbers, our inequalities as presented in \eqref{e:morse inequalities} agree with the classical inequalities \eqref{equ:morse_inequalities} because, in this setting, $\beta_d(X) = p_d$ and $m_d^{\epsilon} = \crit_d(f)$ for $\epsilon > 0$ smaller than the minimal difference between any two critical values of $f$.

The finiteness conditions of our theorem are also satisfied for the functionals that Morse considered in his work on functional topology, with q-tameness being guaranteed by \cref{t:strong local connectedness implies q-tameness} instead of the insufficient condition Morse proposed.
As we have also mentioned before, our inequalities then agree with Morse's historical version from functional topology as presented in \cite{Morse.1940} if the persistent homology of the sublevel set filtration is not only q-tame but also continuous from above, which is satisfied for compact sublevel set filtrations if one uses \v{C}ech homology, as is the case in Morse's work.
Note, however, that in contrast to Morse's version, our inequalities also work without the assumption of continuity from above, expanding their use beyond \v{C}ech type theories.

\begin{rem}
	In addition to the formulation of the inequalities in terms of cap numbers, Morse also proposes a generalized version of critical points, which he calls homotopically critical and which formalizes the idea of a point being critical if it produces topological changes in the sublevel set filtration.
	This notion is employed by Morse and Tompkins in their study of minimal surfaces \cite{Morse.1939}, which we will partly review in \cref{s:surfaces} and for which a thorough historical account may be found in \cite[Section II.6]{Struwe.1988}.
	The usefulness of this notion might be limited though, because for example in the context of Floer theory, critical points of the action functional corresponding to a Hamiltonian usually have infinite index and thus do not produce changes in the topology.
	In this setting, our approach of formulating the inequalities purely in algebraic terms might be more suitable, because while these critical points are not topologically visible, they do correspond to features in the persistence diagram of the filtered Floer homology.
\end{rem}

### 2021-07-26 introduction.tex
Pre-release version

The interplay between the critical set of a function and the topology of its domain is a cornerstone of modern mathematics.
Nowadays, when thinking about the pioneering work of Marston Morse, our first thought probably considers a differentiable function on a closed smooth manifold, but more general settings should also be regarded.
Morse theory in the smooth context was masterfully presented in Milnor's famous book on the subject \cite{MR0163331}, where he also gave a new proof of Bott's periodicity by applying Morse theory to the energy functional of paths in a Riemannian manifold, which notably goes beyond the compact setting.
Another important example of the use of Morse's insights in an infinite context is Floer's work on the Arnold conjecture and its many ramifications in symplectic topology, as surveyed for example in \cite{MR1702944}.
Morse himself worked in a very general setting, publishing in the 1930s a series of papers \cite{Morse.1937, Morse.1938, Morse.1940, MR9102} in which he established the key results of Morse theory in the broad context defined by semi-continuous functionals on metric spaces.
He named this body of work \emph{functional topology} and used it to study questions about minimal surfaces motivated by Douglas' solution to Plateau’s Problem \cite{Douglas.1931}.
In particular, Morse and Tompkins \cite{Morse.1939, Morse.1941} used these techniques to prove the Unstable Minimal Surface Theorem, a result providing conditions for the existence of critical points of the Douglas functional that are not local minima.

While Morse's work on functional topology did not have a long lasting impact on minimal surface theory or the calculus of variations in general, many of his ideas have resurfaced and flourished in other domains.
In particular, in applied topology and topological data analysis, several key insights of Morse have been independently rediscovered as part of the development of \textit{persistent homology}, a technique that provides robust and efficiently computable invariants of filtered spaces using the functorial properties of homology.
Its success in applied topology has motivated a refined abstract theory of persistence that lies on the intersection of geometry, topology, and representation theory.

@@

The primary contribution of this work consists in the introduction of topological conditions for a broad class of filtered spaces ensuring their associated persistence homology modules satisfy Morse inequalities.
As an application, we identify and fix with our conditions a mistake in the proof the Unstable Minimal Surface Theorem of Morse and Tompkins.

>>>>

The interplay between the critical set of a function and the topology of its domain is a cornerstone of modern mathematics.
Nowadays, when thinking about the pioneering work of Marston Morse, our first thought probably considers a differentiable function on a closed smooth manifold, but more general settings should also be regarded.
Morse theory in the smooth context was masterfully presented in Milnor's famous book on the subject \cite{MR0163331}, where he also gave a new proof of Bott's periodicity by applying Morse theory to the energy functional of paths in a Riemannian manifold, which notably goes beyond the compact setting.
Another important example of the use of Morse's insights in an infinite context is Floer's work on the Arnold conjecture and its many ramifications in symplectic topology, as surveyed for example in \cite{MR1702944}.
Morse himself worked in a very general setting, publishing in the 1930s a series of papers \cite{Morse.1937, Morse.1938, Morse.1940} in which he established the key results of Morse theory in the broad context defined by semi-continuous functionals on metric spaces.
He named this body of work \emph{functional topology} and used it to study questions about minimal surfaces motivated by Douglas' solution to Plateau’s Problem \cite{Douglas.1931}.
In particular, Morse and Tompkins \cite{Morse.1939, Morse.1941} used these techniques to prove the Unstable Minimal Surface Theorem, a result providing conditions for the existence of critical points of the Douglas functional that are not local minima.

While Morse's work on functional topology did not have a long lasting impact on minimal surface theory or the calculus of variations in general ---possibly because, as Hildebrandt [4, p. 324] puts it, ``Morse--Tompkins' original paper [is] unreadable and inaccessible for the non-specialist''---
many of his ideas have resurfaced and flourished in other domains.
In particular, in applied topology and topological data analysis, several key insights of Morse have been independently rediscovered as part of the development of \textit{persistent homology}, a technique that provides robust and efficiently computable invariants of filtered spaces using the functorial properties of homology.
Its success in applied topology has motivated a refined abstract theory of persistence that lies on the intersection of geometry, topology, and representation theory.

@@

The primary contribution of this work consists in the introduction of topological conditions on a broad class of filtered spaces ensuring their associated persistent homology modules satisfy Morse inequalities.
As an application, we identify a mistake in the proof the Unstable Minimal Surface Theorem of Morse and Tompkins and fix it with our results.

### 2021-07-26 main.tex
Pre-release version

\institute{
    Ulrich Bauer \at Technical University of Munich \\ \email{mail@ulrich-bauer.org} \and
	Anibal M.~Medina-Mardones \at University of Notre Dame du Lac \at Max Planck Institute for Mathematics in Bonn \\ \email{ammedmar@mpim-bonn.mpg.de} 
	\and
	Maximilian~Schmahl \at Heidelberg University \\ \email{mschmahl@mathi.uni-heidelberg.de}
}

>>>>

\institute{
    Ulrich Bauer \at Technical University of Munich \\ \email{mail@ulrich-bauer.org} \and
	Anibal M.~Medina-Mardones \at Max Planck Institute for Mathematics \at University of Notre Dame du Lac \\ \email{ammedmar@mpim-bonn.mpg.de} 
	\and
	Maximilian~Schmahl \at Heidelberg University \\ \email{mschmahl@mathi.uni-heidelberg.de}
}
